A Hybrid Ensemble Learning Approach to Star-Galaxy Classification

### 2015-05-07 paper/main.tex
update paper.

\documentclass[useAMS,usenatbib]{mn2e}
\usepackage[letterpaper,margin=0.65in]{geometry}
\usepackage[ansinew]{inputenc}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{color}
\usepackage{amssymb}
\usepackage[normalem]{ulem}
\usepackage{natbib}
\usepackage{url}
\usepackage{float}

\def\aap{{ A\&A}}
\def\AaA{{ A\&A}}
\def\AJ{{AJ}}
\def\aj{{AJ}}
\def\ApJ{{ApJ}}
\def\apj{{ApJ}}
\def\apjl{{ApJL}}
\def\ApJL{{ApJL}}
\def\ApJSupp{{ApJS}}
\def\MN{{MNRAS}}
\def\mnras{{MNRAS}}
\def\pasp{{PASP}}
\def\jcap{JCAP}
\def\pasj{{PASJ}}
\def\nat{Nature}
\def\ARAA{{ARA\&A}}
\def\apjs{{ApJS}}
\def\araa{{ARA\&A}}
\def\apss{{Ap\&SS}}
\def\cjaa{{ChJAA}}
\def\aaps{{A\&AS}}

@@

\title{A Hybrid Ensemble Learning Approach to Star-galaxy Classification}

\author[E. J. Kim et al.]{
  Edward J. Kim\thanks{jkim575@illinois.edu}$^1$, Robert J. Brunner$^2$,
  and Matias Carrasco Kind$^3$\\
$^1$Department of Physics, University of Illinois, Urbana, IL 61801 USA\\
$^2$Department of Astronomy, University of Illinois, Urbana, IL 61801 USA\\
$^3$National Center for Supercomputing Applications, Urbana, IL 61801 USA}

@@

\label{firstpage}
\begin{abstract}
\textcolor{blue}{
Star-galaxy separation in currently ongoing and future
large photometric surveys, such as DES and LSST,
will detect a vast number of unresolved galaxies at faint magnitudes.
There exist a variety of star-galaxy classification techniques,
each with their own strenths and weaknesses.
In this paper, we present a novel hybrid, ensemble classification
framework that combines and fully exploits different techniques
to produce a more robust classification.
To demonstrate this hybrid technique,
we combine a purely morphological classifier,
a supervised machine learning technique based on random forest,
an unsupervised method based on self-organizing maps,
and a heirarchical Bayeisan template fitting algorithm.
Any of these four methods could
easily be replaced by a similar method
or be extended to include other existing techniques.
Testing with CFHTLenS data,
we demonstrate that our combination technique improves
the overall performance over any individual classification method.
}

@@

\begin{keywords}
\textcolor{red}{keywords}
\end{keywords}

@@

The problem of star-galaxy classification is fundamental to astronomy
and goes as far back as \cite{messier1781catalogue}.
A variety of different strategies have been developed 
to tackle this long-standing problem,
and yet it remains an unsolved challenge.
The most commonly used method to classify stars and galaxies
in large sky surveys is the morphological separation
~\citep{sebok1979optimal, kron1980photometry, valdes1982resolution,
yee1991faint, vasconcellos2011decision,
henrion2011bayesian}.
It relies on the assumption that
stars appear as point sources
while galaxies appear as resolved sources.
However,
currently ongoing and upcoming large photometric surveys,
such as the Dark Energy Survey
(DES\footnote{http://www.darkenergysurvey.org/})
and the Large Synoptic Survey Telescope
(LSST\footnote{http://www.lsst.org/lsst/}),
will detect a vast number of unresolved galaxies
at faint magnitudes.
Near a survey's limit, the photometric observations
cannot reliably separate stars from unresolved galaxies
by morphology alone without leading to
incompleteness and contamination in the star and galaxy samples.

@@

In supervised learning, 
the input attributes (\eg magnitudes or colors)
are provided along with the truth labels (\eg star or galaxy).
\cite{odewahn1992automated} pioneered
the application of neural networks
to the star-galaxy classification problem,
and it has become a core part of
the astronomical image processing software
\textsc{SExtractor}~\citep{bertin1996sextractor}.
Other successfully implemented examples include
decision trees~\citep{weir1995automated, suchkov2005census, ball2006robust}
and Support Vector Machines~\citep{Fadely2012}.
Unsupervised machine learning techniques
are less common
as they do not utilize the truth labels during the training process
and only the input attributes are used.

@@

In this paper,
we present a novel star-galaxy classification framework
that combines and fully exploits different classification techniques
to produce a more robust classification.
In particular,
we show that the combination of a morphological separation method,
a template fitting technique, a supervised machine learning method,
and an unsupervised machine learning algorithm
can improve the overall performance over any individual method.
In Section~\ref{section:classification_methods},
we describe each of the star-galaxy classification methods.
In Section~\ref{section:classification_combination_methods},
we describe different classification combination techniques
In Section~\ref{section:data},
we describe the Canada-France Hawaii Telescope Lensing Survey (CFHTLenS)
data set with which we test the algorithms.
In Section~\ref{section:results_and_discussion},
we compare the performance of our combination techniques
to the performance of the individual classification techniques.

@@

In this section, we present
four distinct star-galaxy classification techniques.
The first method is
a supervised machine learning technique named \texttt{TPC}
(Trees for Probabilistic Classification),
which uses prediction trees and a random forest~\citep{carrascokind2013tpz}.
The second method is
an unsupervised machine learning technique named \texttt{SOMc},
which uses self-organizing maps (SOM) and a random atlas
to provide a classification~citep{carrascokind2014somz}.
The third method is
a Hierarchical Bayesian template fitting technique
based on the work by \cite{Fadely2012},
which fits SED templates from star and galaxy libraries
to an observed set of measured flux values.
The fourth method is a morphological separation method,
which uses a hard cut in the half-light radius vs.\ magnitude plane.

@@

The simplest and perhaps the most widely used approach
to star-galaxy classification is
to make a hard cut in the space of photometric attributes.
As a first-order morphological selection of point sources,
we adopt a technique that is popular among the weak lensing 
community~\citep{Kaiser1995}.
As Figure~\ref{fig:morph} shows, there is a distinct locus
produced by point sources in the half-light radius
($r_h$; estimated by \textsc{SExtractor}'s \texttt{FLUX\_RADIUS} parameter)
vs.\ the $i$-band magnitude plane.
A rectangular cut in this size-magnitude plane separates point sources,
which are presumed to be stars,
from resolved sources,
which are presumed to be galaxies.
The boundaries of the selection box are determined by
manually inspecting the size-magnitude diagram.

\begin{figure}
  \centering
  \includegraphics[width=0.95\linewidth]{figures/morph.eps}
  \caption{Half-light radius ($r_h$) vs.\ magnitude.}
  \label{fig:morph}
\end{figure}

One of the disadvantages of such cut-based methods is
that it classifies every source with absolute certainty.
It is difficult to justify such a decisive classification
near a survey's magnitude limits,
where measurement uncertainties generally increase.
A more informative approach is to provide probabilistic classifications.
Although a recent work by \cite{henrion2011bayesian}
implemented a probabilistic classification using a Bayesian approach
on the morphological measurements alone,
here we use a cut-based morphological separation
to demonstrate the advantages of our combination techniques.
In particular, we later show that
the binary output (\ie 0 or 1) of cut-based methods
can be transformed into a
probability estimate by combining them
with the probability outputs from other
probabilistic classification techniques,
such as \texttt{TPC}, \texttt{SOMc}, and \texttt{HB}.

@@

\texttt{TPC} is a parallel, supervised machine learning algorithm
that uses prediction trees and random forest 
techniques~\citep{breiman1984classification, breiman2001random}
to produce star-galaxy classification.
\texttt{TPC} is a part of a publicly available software package called
\texttt{MLZ}\footnote{http://lcdm.astro.illinois.edu/code/mlz.html}
(Machine Learning for Photo-$z$).
The full software package includes, among other capabilities,
\texttt{TPZ}, a supervised photo-$z$ technique
\citep[regression mode;][]{carrascokind2013tpz},
\texttt{TPC}, a supervised star-galaxy classification technique
(classification mode),
\texttt{SOM}$z$, an unsupervised photo-$z$ technique
\citep[regression mode;][]{carrascokind2014somz},
and
\texttt{SOMc}, an unsupervised star-galaxy classification technique
(classification mode).

\texttt{TPC} uses classification trees,
a type of prediction trees that are designed to
provide a classification or predict a discrete category.
Prediction trees are built by asking a sequence of questions
that recursively split the data into branches
until a terminal leaf is created
that meets a stopping criterion
(\eg a minimum leaf size).
The optimal split dimension is decided by
choosing the attribute that maximizes
the \textit{Information Gain} ($I_G$), which is defined as

\begin{equation}
  I_G \left(D_{\mathrm{node}}, X\right)
  = I_d \left( D_{\mathrm{node}} \right)
  - \sum_{x \in \mathrm{values}(X)}
  \frac{|D_{\mathrm{node}, x}|}{|D_{\mathrm{node}}|}
  I_d \left( D_{\mathrm{node}, x} \right),
  \label{eq:information_gain}
\end{equation}

\noindent
where $D_{\mathrm{node}}$ is the training data in a given node,
$X$ is one of the possible dimensions (\eg magnitudes or colors)
along which the node is split, and
$x$ are the possible values of a specific dimension $X$.
$|D_{\mathrm{node}}|$ and $|D_{\mathrm{node}, x}|$ are the size of the total training data
and the number of objects in a given subset $x$ within the current node,
respectively.
$I_d$ is the impurity degree index,
and \texttt{TPC} can calculate $I_d$
from any of the three standard different impurity indices:
\textit{information entropy}, \textit{Gini impurity},
and \textit{classification error}.
Here, we use the information entropy,
which is defined similarly to the thermodynamic entropy:

\begin{equation}
  I_d \left( D \right)
  = - f_g \log_{2} f_g - (1 - f_g) \log_{2} f_g,
\end{equation}

@@

In a technique called random forest,
we create bootstrap samples
(\ie $N$ randomly selected objects with replacement)
from the input training data
by sampling repeatedly from the magnitudes using the magnitude errors.
We use these bootstrap samples to construct
multiple, uncorrelated prediction trees
whose individual predictions are aggregated to produce
a star-galaxy classification for each source.

We also use a cross validation technique called
Out-of-Bag~\citep[OOB;][]{breiman1984classification, carrascokind2013tpz}.
When a tree (or a map) is built in \texttt{TPC} (or \texttt{SOMc}),
a fraction of the training data, usually one-third,
is left out and not used in training the trees or maps.
After a tree is constructed using two-thirds of the training data,
the final tree is applied to the remaining one-third
to make a classification.
This process is repeated for every tree,
and the predictions from each tree are aggregated
for each object to make the final star-galaxy classification.
We emphasize that if an object is used for training a given tree,
it is never used for subsequent prediction by that tree.
Thus, the OOB data is an unbiased estimation of the errors
and can be used as cross-validation data
as long as the OOB data remain similar to the final test data set.
The OOB technique can also provide extra information such as
a ranking of the relative importance of the input attributes
used in the prediction.
The OOB technique can prove extremely valuable
when calibrating the algorithm,
when deciding which attributes to incorporate in the construction of the trees,
and when combining this approach with other techniques.

@@

A Self-Organizing Map
\citep[SOM;][]{kohonen1990self, kohonen2001self}
is an unsupervised, artificial neural network algorithm
that is capable of projecting high-dimensional input data
onto a low-dimensional map
through a process of competitive learning.
In astronomical applications,
the high-dimensional input data can be
magnitudes, colors, or some other photometric attributes.
The output map is usually chosen to be two-dimensional
for the visualization.
The differences between a SOM and other neural network algorithms are
that a SOM is unsupervised,
there are no hidden layers and therefore no extra parameters,
and it produces a direct mapping
between the training set and the output network.
In fact, a SOM can be viewed as a non-linear generalization
of a principal component analysis (PCA) algorithm.

@@

In a similar approach to random forest in \texttt{TPZ} and \texttt{TPC},
\texttt{SOM}$z$ uses a technique called \textit{random atlas}
to provide photo-$z$ estimation \citep{carrascokind2014somz}.
In random atlas, the prediction trees of random forest are replaced by maps,
and each map is constructed from different bootstrap samples
of the training data.
Furthermore, we create random realizations of the training data
by perturbing the input attributes by their measurement error.
For each map, we can either use all available attributes,
or randomly select a subsample of the attribute space.
This SOM implementation can also be
applied to the classification problem,
and we refer to it as \texttt{SOMc}
in order to differentiate it from 
the photo-$z$ estimation problem (regression mode).
We also use the random atlas approach
in some of the classification combination approaches as discussed in
Section~\ref{section:classification_combination_methods}.

One of the most important parameter in \texttt{SOMc}
is the topology of the two-dimensional SOM,
which can be rectangular, hexagonal, or spherical.
To classify stars and galaxies in the CFHTLenS data,
we use a spherical topology,
which is constructed by using
\texttt{HEALPIX}~\citep{gorski2005healpix}.
Furthermore, similar to \texttt{TPC},
we use the OOB technique to make an unbiased estimation of errors.
For a complete description of the SOM implementation
and its application to the estimation of
photo-$z$ probability distribution functions,
we refer the reader to \cite{carrascokind2014somz}.

@@

\begin{equation}
P \left( S | \mathbf{x}, \mathbf{\theta} \right)
= P \left( \mathbf{x} | S, \mathbf{\theta} \right)
P \left( S | \mathbf{\theta} \right),
\end{equation}

\noindent
where $\mathbf{x}$ represents a given set of observed magnitudes,.
We have also introduced the \textit{hyperparameter} $\mathbf{\theta}$,
a nuisance parameter that characterizes our uncertainty
in the prior distribution.
To compute the likelihood that a source is a star,
we marginalize over all star and galaxy templates $\mathbf{T}$.
In a template-fitting approach,
we marginalize by summing up
the likelihood that a source has the set of magnitudes $\mathbf{x}$
for a given star template
as well as the likelihood for a given galaxy template:

\begin{equation}
  P \left(\mathbf{x} | S, \mathbf{\theta} \right)
  = \sum_{t \in \mathbf{T}}
  P \left(\mathbf{x} | S, t, \mathbf{\theta} \right)
  P \left(t | S, \mathbf{\theta} \right).
  \label{eq:marginalize_template}
\end{equation}

\noindent
The likelihood of each template
$P \left( \mathbf{x} | S, \mathbf{\theta} \right)$
is itself marginalized over the uncertainty
in the template-fitting coefficient.
Furthermore, for galaxy templates, we introduce another step that 
marginalizes the likelihood by redshifting a given galaxy template
by a factor of $1 + z$.

Marginalization in Equation~\ref{eq:marginalize_template}
requires that we specify the prior probability
$P \left(t | S, \mathbf{\theta} \right)$
that a source has spectral template $t$ (at a given redshift).
Thus, the probability that a source is a star (or a galaxy)
is either the posterior probability itself
if a prior is used,
or the likelihood itself if an uniformative prior is used.
In a Bayesian analysis, it is preferable to use a prior,
which can be directly computed either from physical assumptions,
from an empirical function calibrated by
using a spectroscopic training sample,
or from an empirical function calibrated by
using machine learning techniques.
In an HB approach, the entire sample of sources is used to
infer the prior probabilities for each individual source.

@@

\begin{equation}
\sum_{t \in \mathbf{T}}
P \left(t | S, \mathbf{\theta} \right) = 1.
\end{equation}

\noindent
These weights correspond to the hyperparameters,
which can be inferred by sampling
the posterior probability distribution in the hyperparameter space.
For the sampling, we use \texttt{emcee}, a Python implementation of the
affine-invariant Markov Chain Monte Carlo (MCMC) ensemble sampler~\citep{Foreman-Mackey2013}.

As the goal of template fitting methods is to minimize
the difference between observed and theoretical magnitudes,
this approach heavily relies on
both the use of SED templates and
the accuracy of the transmission functions
for the filters used for particular survey.
For our stellar templates,
we use the empirical SED library from \cite{pickles1998stellar}.
The Pickles library consists of 131 stellar templates,
which span all normal spectral types
and luminosity classes at solar abundance,
as well as metal-poor and metal-rich F-K dwarf 
and G-K giant and supergiant stars.
\textcolor{blue}{
We supplement the stellar library with
100 SEDs from \cite{chabrier2000evolutionary},
which include low mass stars and brown dwarfs
with different $T_{eff}$ and surface gravities.
We also include four white dwarf templates of
\cite{bohlin1995white}.
}
For our galaxy templates,
we use the four CWW spectra from \cite{coleman1980colors},
which include an Elliptical, an Sba, an Sbb,
and an Irregular galaxy template.
When extending an analysis to higher redshift,
the CWW library is often augmented with
two star bursting galaxy templates from \cite{kinney1996template}.
\textcolor{blue}{
From the six original CWW and Kinney spectra,
intermediate templates are created by interpolation
for a total of 51 SEDs.
}

@@

Building on the work in the field of ensemble learning,
we combine the predictions from
individual star-galaxy classification techniques
using ensemble learning techniques known as
Bayesian Model Combination (BMC)
and a variant of an ensemble learning technique known as stacking.
The main idea behind ensemble learning is to weigh
the predictions from individual models
and combine them to obtain a prediction
that outperforms every one of 
them individually~\citep{rokach2010ensemble}.

@@

Given the variety of star-galaxy classification methods
we are using,
we fully expect the relative performance
of the individual techniques to vary across
the parameter space spanned by the data.
For example, it is reasonable to expect 
supervised techniques to outperform other techniques
in areas of parameter space that are well-populated
with training data.
Similarly, we can expect unsupervised approaches
such as \texttt{SOM} or template fitting approaches 
to generally perform better when training data
is either sparse or unavailable.

>>>>

\documentclass[useAMS,usenatbib]{mn2e}
\usepackage[letterpaper,margin=0.65in]{geometry}
\usepackage[ansinew]{inputenc}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{color}
\usepackage{amssymb}
\usepackage[normalem]{ulem}
\usepackage{natbib}
\usepackage{url}
\usepackage{placeins}
\usepackage{times}

% By default the URLs are put in typewriter type in the body and the
% bibliography of the document when using the \url command.  If you are
% using many long URLs you may want to uncommennt the next line so they
% are typeset a little smaller.
\renewcommand{\UrlFont}{\small\tt}

\def\aap{{A\&A}}
\def\AaA{{A\&A}}
\def\AJ{{AJ}}
\def\aj{{AJ}}
\def\ApJ{{ApJ}}
\def\apj{{ApJ}}
\def\apjl{{ApJL}}
\def\ApJL{{ApJL}}
\def\ApJSupp{{ApJS}}
\def\MN{{MNRAS}}
\def\mnras{{MNRAS}}
\def\pasp{{PASP}}
\def\jcap{JCAP}
\def\pasj{{PASJ}}
\def\nat{Nature}
\def\ARAA{{ARA\&A}}
\def\apjs{{ApJS}}
\def\araa{{ARA\&A}}
\def\apss{{Ap\&SS}}
\def\cjaa{{ChJAA}}
\def\aaps{{A\&AS}}

@@

\pdfminorversion=5

\title{A Hybrid Ensemble Learning Approach to Star-Galaxy Classification}

\author[E. J. Kim, R. J. Brunner \& M. Carrasco Kind]{
  Edward J.~Kim$^1$\thanks{jkim575@illinois.edu}, Robert J.~Brunner$^2$,
  and Matias Carrasco Kind$^{2,3}$\\
$^1$Department of Physics, University of Illinois, Urbana, IL 61801 USA\\
$^2$Department of Astronomy, University of Illinois, Urbana, IL 61801 USA\\
$^3$National Center for Supercomputing Applications, Urbana, IL 61801 USA}

@@

\label{firstpage}
\begin{abstract}
%Star-galaxy separation in currently ongoing and future
%large photometric surveys, such as DES and LSST,
%will detect a vast number of unresolved galaxies at faint magnitudes.
There exist a variety of star-galaxy classification techniques,
each with their own strengths and weaknesses.
In this paper, we present a novel meta-classification
framework that combines and fully exploits different techniques
to produce a more robust star-galaxy classification.
To demonstrate this hybrid, ensemble approach,
we combine a purely morphological classifier,
a supervised machine learning method based on random forest,
an unsupervised machine learning method based on self-organizing maps,
and a Hierarchical Bayesian template fitting method.
%Any of these four methods could
%easily be replaced by a similar method
%or be extended to include other existing techniques.
Using data from the CFHTLenS survey,
we consider different scenarios:
when a high-quality training set is available with spectroscopic labels from
DEEP2, SDSS, VIPERS, and VVDS, and
when the demographics of sources in a low-quality training set
do not match the demographics of objects in the test data set.
We demonstrate that our Bayesian combination technique improves
the overall performance over any individual classification method
in these scenarios.
Thus, strategies that combine the predictions of different classifiers
may prove to be optimal in currently ongoing and forthcoming
photometric surveys,
such as the Dark Energy Survey and the Large Synoptic Survey Telescope.

@@

\begin{keywords}
methods: data analysis -- methods: statistical -- surveys -- stars: statistics
-- galaxies:statistics.
\end{keywords}

@@

The problem of source classification is fundamental to astronomy
and goes as far back as \citet{messier1781catalogue}.
A variety of different strategies have been developed 
to tackle this long-standing problem,
%and yet it remains an unsolved challenge.
and yet there is no consensus on
the optimal star-galaxy classification strategy.
The most commonly used method to classify stars and galaxies
in large sky surveys is the morphological separation
~\citep{sebok1979optimal, kron1980photometry, valdes1982resolution,
yee1991faint, vasconcellos2011decision,
henrion2011bayesian}.
It relies on the assumption that
stars appear as point sources
while galaxies appear as resolved sources.
However,
currently ongoing and upcoming large photometric surveys,
such as the Dark Energy Survey
(DES\footnote{http://www.darkenergysurvey.org/})
and the Large Synoptic Survey Telescope
(LSST\footnote{http://www.lsst.org/lsst/}),
will detect a vast number of unresolved galaxies
at faint magnitudes.
Near a survey's limit, the photometric observations
cannot reliably separate stars from unresolved galaxies
by morphology alone without leading to
incompleteness and contamination in the star and galaxy samples.

@@

In supervised learning, 
the input attributes (\eg magnitudes or colors)
are provided along with the truth labels (\eg star or galaxy).
\citet{odewahn1992automated} pioneered
the application of neural networks
to the star-galaxy classification problem,
and it has become a core part of
the astronomical image processing software
\textsc{SExtractor}~\citep{bertin1996sextractor}.
Other successfully implemented examples include
decision trees~\citep{weir1995automated, suchkov2005census, ball2006robust,
sevilla2015effect}
and Support Vector Machines~\citep*{Fadely2012}.
Unsupervised machine learning techniques
are less common,
as they do not utilize the truth labels during the training process,
and only the input attributes are used.

@@

In this paper,
we present a novel star-galaxy classification framework
that combines and fully exploits different classification techniques
to produce a more robust classification.
In particular,
we show that the combination of a morphological separation method,
a template fitting technique, a supervised machine learning method,
and an unsupervised machine learning algorithm
can improve the overall performance over any individual method.
In Section~\ref{section:classification_methods},
we describe each of the star-galaxy classification methods.
In Section~\ref{section:classification_combination_methods},
we describe different classification combination techniques
In Section~\ref{section:data},
we describe the Canada-France Hawaii Telescope Lensing Survey (CFHTLenS)
data set with which we test the algorithms.
In Section~\ref{section:results_and_discussion},
we compare the performance of our combination techniques
to the performance of the individual classification techniques.
Finally, we outline our conclusions in Section~\ref{section:conclusions}.

@@

In this section, we present
four distinct star-galaxy classification techniques.
The first method is a morphological separation method,
which uses a hard cut in the half-light radius vs.\ magnitude plane.
The second method is
a supervised machine learning technique named TPC
(Trees for Probabilistic Classification),
which uses prediction trees and a random forest~\citep{carrascokind2013tpz}.
The third method is
an unsupervised machine learning technique named SOMc,
which uses self-organizing maps (SOMs) and a random atlas
to provide a classification~\citep{carrascokind2014somz}.
The fourth method is
a Hierarchical Bayesian template fitting technique
based on the work by \citet{Fadely2012},
which fits SED templates from star and galaxy libraries
to an observed set of measured flux values.

@@

The simplest and perhaps the most widely used approach
to star-galaxy classification is
to make a hard cut in the space of photometric attributes.
As a first-order morphological selection of point sources,
we adopt a technique that is popular among the weak lensing 
community~\citep*{Kaiser1995}.
As Figure~\ref{fig:morph} shows, there is a distinct locus
produced by point sources in the half-light radius
(estimated by \textsc{SExtractor}'s FLUX\_RADIUS parameter)
vs.\ the $i$-band magnitude plane.
A rectangular cut in this size-magnitude plane separates point sources,
which are presumed to be stars,
from resolved sources,
which are presumed to be galaxies.
The boundaries of the selection box are determined by
manually inspecting the size-magnitude diagram.

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{figures/morph.eps}
  \caption{Half-light radius vs.\ magnitude.}
  \label{fig:morph}
\end{figure}

One of the disadvantages of such cut-based methods is
that it classifies every source with absolute certainty.
It is difficult to justify such a decisive classification
near a survey's magnitude limits,
where measurement uncertainties generally increase.
A more informative approach is to provide probabilistic classifications.
Although a recent work by \citet{henrion2011bayesian}
implemented a probabilistic classification using a Bayesian approach
on the morphological measurements alone,
here we use a cut-based morphological separation
to demonstrate the advantages of our combination techniques.
In particular, we later show that
the binary outputs (\ie 0 or 1) of cut-based methods
can be transformed into
probability estimates by combining them
with the probability outputs from other
probabilistic classification techniques,
such as TPC, SOMc, and HB.

@@

TPC is a parallel, supervised machine learning algorithm
that uses prediction trees and random forest 
techniques~\citep{breiman1984classification, breiman2001random}
to produce a star-galaxy classification.
TPC is a part of a publicly available software package called
\textsc{MLZ}\footnote{http://lcdm.astro.illinois.edu/code/mlz.html}
(Machine Learning for Photo-$z$).
The full software package includes:
\textsc{TPZ}, a supervised photometric redshift (photo-$z$)
estimation technique
\citep[regression mode;][]{carrascokind2013tpz};
TPC, a supervised star-galaxy classification technique
(classification mode);
\textsc{SOM}$z$, an unsupervised photo-$z$ technique
\citep[regression mode;][]{carrascokind2014somz};
and
SOMc, an unsupervised star-galaxy classification technique
(classification mode).

TPC uses classification trees,
a type of prediction trees that are designed to
provide a classification or predict a discrete category.
Prediction trees are built by asking a sequence of questions
that recursively split the data into branches
until a terminal leaf is created
that meets a stopping criterion
(\eg a minimum leaf size).
The optimal split dimension is decided by
choosing the attribute that maximizes
the \textit{Information Gain} ($I_G$), which is defined as

\begin{equation} \label{eq:information_gain}
  I_G \left(D_{\rmn{node}}, X\right)
  = I_d \left( D_{\rmn{node}} \right)
  - \sum_{x \in \rmn{values}(X)}
  \frac{|D_{\rmn{node}, x}|}{|D_{\rmn{node}}|}
  I_d \left( D_{\rmn{node}, x} \right),
\end{equation}

\noindent
where $D_{\rmn{node}}$ is the training data in a given node,
$X$ is one of the possible dimensions (\eg magnitudes or colors)
along which the node is split, and
$x$ are the possible values of a specific dimension $X$.
$|D_{\rmn{node}}|$ and $|D_{\rmn{node}, x}|$ are the size of the total training data
and the number of objects in a given subset $x$ within the current node,
respectively.
$I_d$ is the impurity degree index,
and TPC can calculate $I_d$
from any of the three standard different impurity indices:
\textit{information entropy}, \textit{Gini impurity},
and \textit{classification error}.
In this work, we use the information entropy,
which is defined similarly to the thermodynamic entropy:

\begin{equation}
  I_d \left( D \right)
  = - f_g \log_{2} f_g - \left(1 - f_g\right) \log_{2} \left(1 - f_g\right),
\end{equation}

@@

In a technique called random forest,
we create bootstrap samples
(\ie $N$ randomly selected objects with replacement)
from the input training data
by sampling repeatedly from the magnitudes and colors
using their measurement errors.
We use these bootstrap samples to construct
multiple, uncorrelated prediction trees
whose individual predictions are aggregated to produce
a star-galaxy classification for each source.

We also use a cross-validation technique called
Out-of-Bag~\citep[OOB;][]{breiman1984classification, carrascokind2013tpz}.
When a tree (or a map) is built in TPC (or SOMc),
a fraction of the training data, usually one-third,
is left out and not used in training the trees (or maps).
After a tree is constructed using two-thirds of the training data,
the final tree is applied to the remaining one-third
to make a classification.
This process is repeated for every tree,
and the predictions from each tree are aggregated
for each object to make the final star-galaxy classification.
We emphasize that if an object is used for training a given tree,
it is never used for subsequent prediction by that tree.
Thus, the OOB data is an unbiased estimation of the errors
and can be used as cross-validation data
as long as the OOB data remain similar to the final test data set.
The OOB technique can also provide extra information such as
a ranking of the relative importance of the input attributes
used in the prediction.
The OOB technique can prove extremely valuable
when calibrating the algorithm,
when deciding which attributes to incorporate in the construction of the trees,
and when combining this approach with other techniques.

@@

A self-organizing map~\citep{kohonen1990self, kohonen2001self}
is an unsupervised, artificial neural network algorithm
that is capable of projecting high-dimensional input data
onto a low-dimensional map
through a process of competitive learning.
In astronomical applications,
the high-dimensional input data can be
magnitudes, colors, or some other photometric attributes.
The output map is usually chosen to be two-dimensional
so that the resulting map can be used for visualizing
various properties of the input data.
The differences between a SOM and other neural network algorithms are
that a SOM is unsupervised,
there are no hidden layers and therefore no extra parameters,
and it produces a direct mapping
between the training set and the output network.
In fact, a SOM can be viewed as a non-linear generalization
of a principal component analysis (PCA) algorithm~\citep{yin2008self}.

@@

In a similar approach to random forest in TPZ and TPC,
SOM$z$ uses a technique called \textit{random atlas}
to provide photo-$z$ estimation \citep{carrascokind2014somz}.
In random atlas, the prediction trees of random forest are replaced by maps,
and each map is constructed from different bootstrap samples
of the training data.
Furthermore, we create random realizations of the training data
by perturbing the magnitudes and colors by their measurement errors.
For each map, we can either use all available attributes,
or randomly select a subsample of the attribute space.
This SOM implementation can also be
applied to the classification problem,
and we refer to it as SOMc
in order to differentiate it from 
the photo-$z$ estimation problem (regression mode).
We also use the random atlas approach
in some of the classification combination approaches as discussed in
Section~\ref{section:classification_combination_methods}.

One of the most important parameter in SOMc
is the topology of the two-dimensional SOM,
which can be rectangular, hexagonal, or spherical.
To classify stars and galaxies in the CFHTLenS data,
we use a spherical topology,
which is constructed by using
\textsc{HEALPIX}~\citep{gorski2005healpix}.
Furthermore, similar to TPC,
we use the OOB technique to make an unbiased estimation of errors.
For a complete description of the SOM implementation
and its application to the estimation of
photo-$z$ probability density functions (photo-$z$ PDFs),
we refer the reader to \cite{carrascokind2014somz}.

@@

\begin{equation} \label{eq:overall_posterior}
P \left( S | \bmath{x}, \bmath{\theta} \right)
= P \left( \bmath{x} | S, \bmath{\theta} \right)
P \left( S | \bmath{\theta} \right),
\end{equation}

\noindent
where $\bmath{x}$ represents a given set of observed magnitudes,.
We have also introduced the \textit{hyperparameter} $\bmath{\theta}$,
a nuisance parameter that characterizes our uncertainty
in the prior distribution.
To compute the likelihood that a source is a star,
we marginalize over all star and galaxy templates $\bmath{T}$.
In a template-fitting approach,
we marginalize by summing up
the likelihood that a source has the set of magnitudes $\bmath{x}$
for a given star template
as well as the likelihood for a given galaxy template:

\begin{equation} \label{eq:marginalize_template}
  P \left(\bmath{x} | S, \bmath{\theta} \right)
  = \sum_{t \in \bmath{T}}
  P \left(\bmath{x} | S, t, \bmath{\theta} \right)
  P \left(t | S, \bmath{\theta} \right).
\end{equation}

\noindent
The likelihood of each template
$P \left( \bmath{x} | S, \bmath{\theta} \right)$
is itself marginalized over the uncertainty
in the template-fitting coefficient.
Furthermore, for galaxy templates, we introduce another step that 
marginalizes the likelihood by redshifting a given galaxy template
by a factor of $1 + z$.

Marginalization in Equation~\ref{eq:marginalize_template}
requires that we specify the prior probability
$P \left(t | S, \bmath{\theta} \right)$
that a source has a spectral template $t$ (at a given redshift).
Thus, the probability that a source is a star (or a galaxy)
is either the posterior probability itself
if a prior is used,
or the likelihood itself if an uninformative prior is used.
In a Bayesian analysis, it is preferable to use a prior,
which can be directly computed either from physical assumptions,
or from an empirical function calibrated by
using a spectroscopic training sample.
In an HB approach, the entire sample of sources is used to
infer the prior probabilities for each individual source.

@@

\begin{equation}
\sum_{t \in \bmath{T}}
P \left(t | S, \bmath{\theta} \right) = 1.
\end{equation}

\noindent
Similarly, we also parametrize the overall prior probability,
$\left(S | \bmath{\theta}\right)$ in Equation~\ref{eq:overall_posterior},
as a weight.
These weights correspond to the hyperparameters,
which can be inferred by sampling
the posterior probability distribution in the hyperparameter space.
For the sampling, we use \textsc{emcee}, a Python implementation of the
affine-invariant Markov Chain Monte Carlo (MCMC) ensemble sampler~\citep{Foreman-Mackey2013}.

As the goal of template fitting methods is to minimize
the difference between observed and theoretical magnitudes,
this approach heavily relies on
both the use of SED templates and
the accuracy of the transmission functions
for the filters used for particular survey.
For our stellar templates,
we use the empirical SED library from \citet{pickles1998stellar}.
The Pickles library consists of 131 stellar templates,
which span all normal spectral types
and luminosity classes at solar abundance,
as well as metal-poor and metal-rich F--K dwarf 
and G--K giant and supergiant stars.
We supplement the stellar library with
100 SEDs from \citet{chabrier2000evolutionary},
which include low mass stars and brown dwarfs
with different $T_{\rmn{eff}}$ and surface gravities.
We also include four white dwarf templates of
\citet*{bohlin1995white}, for a total of 235 templates
in our final stellar library.  
For our galaxy templates,
we use four CWW spectra from \cite*{coleman1980colors},
which include an Elliptical, an Sba, an Sbb,
and an Irregular galaxy template.
When extending an analysis to higher redshifts,
the CWW library is often augmented with
two star bursting galaxy templates from \cite{kinney1996template}.
From the six original CWW and Kinney spectra,
intermediate templates are created by interpolation,
for a total of 51 SEDs in our final galaxy library.

@@

Building on the work in the field of ensemble learning,
we combine the predictions from
individual star-galaxy classification techniques
using four combination techniques,
including Bayesian Model Combination (BMC)
and a variant of an ensemble learning technique known as stacking.
The main idea behind ensemble learning is to weight
the predictions from individual models
and combine them to obtain a prediction
that outperforms every one of 
them individually~\citep{rokach2010ensemble}.

@@

Given the variety of star-galaxy classification methods
we are using,
we fully expect the relative performance
of the individual techniques to vary across
the parameter space spanned by the data.
For example, it is reasonable to expect 
supervised techniques to outperform other techniques
in areas of parameter space that are well-populated
with training data.
Similarly, we can expect unsupervised approaches
such as SOM or template fitting approaches 
to generally perform better when training data
is either sparse or unavailable.
